---
title: "Socmixmods"
author: "Samuel Ellis"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Socmixmods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

Socmixmods is a package designed to facilitate researchers applying binomial mixture models to social data. The use of binomial mixture models in social data was introduced in the paper: Weiss MN, Franks DW, Croft DP, Whitehead H (2019) *Measuring the complexity of social associations using mixture models.* Behav Ecol Sociobiol (https://doi.org/10.1007/s00265-018-2603-6). This package is an implementation of the methods developed in this paper, indeed it is partly based upon the R code published alongside the paper. A detailed description of the methods and the motivation for using them can be found in this publication. In addition functionality introduced in [Ellis et al upcoming paper] is also included, and again a detailed description of those methods and their application can be found in that publication. This document is intended as a brief guide to how to apply the package to social data.

## Installing socmixmods
Socmixmods is available on github at : https://github.com/samellisq/socmixmods. 

To download from github I recommend the functionality in the devtools package but other methods will also work. 

```{r, eval = FALSE}
install.packages("devtools") # install devtools if not already installed (also avaliable on CRAN)
devtools::install_github("samellisq/socmixmods")

library(socmixmods)

```

## Data

In a binomial mixture model of social data we assume that the number of times two individuals (*i* and *j*) are seen together (x~ij~) is the result of the number of times both members of the pair are observed (d~ij~) and their real social preference *mu*~ij~. Hence:  x~ij~ ~ binomial(d~ij~, *mu*~ij~). In social network terms *x* and *d* represent the numerator and demonator (respectively) of the commonly used Simple Ratio Index. 

The input data for the binomial mixture models is therefore a numeric vector of the denominator and numerator of the Simple Ratio Index. 

```{r, echo = FALSE, message=FALSE}
devtools::load_all()

dat = socmixmod_simulate_data(K = 4, mean.d = 120)
x = dat$n
d = dat$d
edge.ids = dat$edge.ids

```
```{r}
head(data.frame(x, d))
```

In addition for analysis aiming to classifying edges into components (or other edge based analyses) it is useful to have a vector of names for the edges to faciliate linking the output back to data. These edge ids can either be a vector of edge names (or a key value) or a two column matrix of names of the nodes being connected by the edge. If edge ids are not supplied edges are named sequentially. 

```{r}
head(data.frame(x, d, edge.ids))
```

NB: data used here and throughout are simulated via the socmixmod_simulate_data() function explained below. 

## Fitting a model: the **fit_binom_mixture_model** function

The fit_binom_mixture_model function fits a binomial mixture model of K components
to the data. 

```{r}
model = fit_binom_mixture_model(Den = d,
                                Num = x, 
                                J = 3, 
                                edge.ids = edge.ids 
                                )
```

A summary of the model can be accessed with:

```{r}
summary(model)
```

And the edge data, the probability that each edge belongs in each component, can be accessed with:
```{r}
edge.data = by_edge(model)
head(edge.data)
```

Summary information and edge data can be accessed by treating the output as a list

```{r}
head(model$mu) # for edge data
```


Lastly, a plot showing how the components fit onto the social association Simple Ratio Index can be accessed with (note ggplot2 package required):

```{r}
plot(model)
```

Please note that this plot is just for guidance as the model does not actually fit to the Simple Ratio Index rather it fits to the data itself. 

## How many componenents?: the *binom_assoc_mixt* function

Researchers will often want to select the number of components that best fits the data. Socmixmods includes the binom_assoc_mixt function to facilitate this. 

```{r}
model.fitting = binom_assoc_mixt(Den = d, 
                                 Num = x, 
                                 edge.ids = edge.ids, 
                                 minJ = 1,
                                 maxJ = 5,
                                 criterion = "AIC",
                                 run.all.J = TRUE,
                                 verbose = FALSE
                                )
```

This function will the fit, in this case, 5 binomial mixture models to the data: from minJ to
maxJ components. The *run.all.J* parameter specifies whether to run all the models with between minJ and maxJ components (when *TRUE*), or to stop running the model when 2 models in a row have an increasing *criterion* (when *FALSE*). This is particualry pertinent because models with a large number of components can be slow to fit. 

A summary of all 5 models can be accessed via:

```{r}
summary(model.fitting)
```

Each model can be individually viewed by using the $all.models. Each model is produced by the fit_binom_mixture_model function and the same manipulations as described above can be applied:

```{r}
summary(model.fitting$allmodels.list[[4]])
```

And the best fitting model, selected by *criterion*, can be extracted directly with:

```{r}
best.model = get_bestmodel(model.fitting, verbose = FALSE)
summary(best.model)
```




